---
title: "Simulating data Experiment with nSubject"
output: html_notebook
---
## Simulations for nnp experiment (adapted from laurel brehm, feb 11 2018)

```{r packages}
library(lme4)
library(plyr)
```

We generate some data based on information from literature (Liu et al., 2016).
There are 6 conditions because we have a 2 x 3 factorial design (within-subject and within-item design). Specifically, we have GroupMembership (ingroup|outgroup) x TrialType (neutral|invalid|valid).

Basically, we will vary s for subs and i for items. Let's set the variables.
```{r set the variables}
s= 40 #let's have 40 participants
ir= 216  #how many times is each item repeated = trials per item
i=2*ir ## 2 items (the two faces)-- repeated 216times each in the whole experiment
si=s*i ## total observations in study
```

To know the number of total observations per condition, per participant: (si/6)/s.

### Draw random samples for each variable around effect from literature
```{r drawing samples}
#These create the raw values = OUR DEPENDENT MEASURE -->Reaction times
ininv <-rnorm(si/6, mean=685, sd=79)
inval <-rnorm(si/6, mean=719, sd=63)
inneu <-rnorm(si/6, mean=712, sd=91)
outinv <-rnorm(si/6, mean=694, sd=84)
outval <-rnorm(si/6, mean=685, sd=79)
outneu <-rnorm(si/6, mean=719, sd=63)

#Set up vars for subs and items
#this line of code creates the Subject ID vector 1x si (total number of observation in the study)
#and repeats each subject id for the number of trial per condition, multiplied by conditon number 
SubjectID <- rep(rep(1:s, each=i/6),6)  ## subjects per condition: everyone sees all trials

#this line of code creates the item ID vector 1x si (total number of observation in the study)
#and repeats each item id for the number of trial per condition
itemID <- rep(1:6,si/6)
#the two vectors should have same length
```

### Create dataframe and export

```{r dataframe}
## create data frame with all of them
RT <- c(ininv,inval,inneu,outinv,outval,outneu)
Group_Membership <- rep(c(rep("ingroup", si/144),rep("outgroup",si/144)),72)
Trial_type <- rep( c(rep("invalid", si/432),rep("valid",si/432),rep("neutral",si/432)),144)
ds <- cbind(RT,Group_Membership,Trial_type,SubjectID,itemID)
ds <- as.data.frame(ds)
ds$RT <- as.numeric(as.character(ds$RT))
#write
write.csv(ds, "SimulatedData.csv")
```






