---
title: "Solution Data Cleaning"
author: "Giovanna de Vincenzo"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output:
  pdf_document:
    toc: yes
  html_document:
    fig_caption: yes
    theme: flatly
    toc: yes
---

## Load required packages

```{r library, message = FALSE, warning = FALSE}
library(readr)
library(dplyr)
library(stringr)
library(reshape)
```

## Import dataset

```{r dataset_import, eval=TRUE}
markets <- read.csv("Farmers_Markets.csv")
class(markets) 
```

## Inspect your data

First six rows of dataset
```{r head}
head(markets)
```

Variable names
```{r names}
names(markets)
```

Dimensions of data frame
```{r dim}
dim(markets)
```

How many NAs do we have per variable
```{r summary}
summary(markets)
```

Data-viewer
```{r view}
View(markets)
```

Display internal structure
```{r str}
str(markets) 
```

## Dealing with missing values

```{r miss val, message = FALSE, warning = FALSE}

#replace blanks/spaces with NAs
markets[markets==""| markets==" "] = NA

#for factor variables:
markets <- markets %>%
  mutate_if(is.factor, funs(factor(replace(., .=="" | .==" ", NA))))

#Question: What do you do when there are text NAs where actual NAs should be?

markets[markets=="NA"]<-NA

```

We can solve all of the problems above by changing the arguments when importing data:
```{r change}
markets <- read.csv("Farmers_Markets.csv", na.strings=c("NA","NaN", " ", ""))
str(markets)
```

## Variable types

Find out what class each variable in the dataset is
```{r variable_type}
str(markets)
#could also use 
#lapply(markets, class)
```

why don't we have any character variables?
default setting for read.csv is to import strings as factors
to change this, and reimport the dataset with strings as characters we could run:

```{r new import}
markets <- read.csv("Farmers_Markets.csv", stringsAsFactors = F, na.strings=c("NA","NaN", " ", ""))
str(markets)
#now all variables containing strings are of class 'character'
```

## Change class of variable

What if we want to consider a character variable as a factor?
In this dataset, "NEIGHBORHOOD" is the most likely factor variable since it has discernable levels
```{r class_change}
markets$NEIGHBORHOOD <- as.factor(markets$NEIGHBORHOOD)
class(markets$NEIGHBORHOOD) # now it is a factor
```

Factor variables have levels. Notice that some of these levels are redundant

```{r levels}
levels(markets$NEIGHBORHOOD)
```

## Dealing with inconsistent values
White space in variable values

https://bookdown.org/lyzhang10/lzhang_r_tips_book/how-to-deal-with-empty-spaces.html

```{r help}
help(str_trim)

markets <- markets %>% #note this saves the changes to the dataframe
  mutate(NEIGHBORHOOD = str_trim(NEIGHBORHOOD))
```

```{r unique}
unique(markets$NEIGHBORHOOD)
```

## Capitalization in variable values using grep

```{r capitalization}
index <- grepl("city", markets$NEIGHBORHOOD, ignore.case = TRUE) #CITY cITY cItY
```

```{r new_cap}
markets$NEIGHBORHOOD[index] = "Center City"
```

```{r unique_verify}
unique(markets$NEIGHBORHOOD)
```

Let's check our work

```{r check}
levels(markets$NEIGHBORHOOD)
```

Why is this back to being a character? str_trim and grepl operate on character vectors so they likely coerce NEIGHBORHOOD to character
```{r charToFactor}
class(markets$NEIGHBORHOOD)

markets$NEIGHBORHOOD <- as.factor(markets$NEIGHBORHOOD) #now we are back to meaningful levels

```


## Reshaping / pivoting data
Right now our data is in 'long' format, if we wanted to organize by NEIGHBORHOOD and put the data in 'wide' format, we could use reshape() in Base R:
```{r reshaping}
markets_wide <- reshape(markets, idvar = "NEIGHBORHOOD", timevar = "OBJECTID", direction = "wide")
View(markets_wide)
```

To pivot data, we can use the reshape package and the melt and cast functions

## Saving out your cleaned dataset

This is useful so you don't need to repeat each of these steps for future analysis!
Save the script you clean your data in, and save a copy of the raw data for reference

```{r saving}
write.csv(markets, "cleaned_Farmers_Markets.csv")

```
