---
title: "RLadies_Giugno"
author: "Francesca Lamontanara"
date: "20 giugno 2019"
output: html_document
---

##Data Visualization with ggplot2
###R-Ladies:25/06/2019    

```{r}
#Load packages
library(ggplot2)
library(dplyr)
```


```{r}
mpg

```
Variables are:

* manufacturer

* model= model name

* displ= engine displacement, in litres

* year= year of manufacture

* cyl= number of cylinders

* trans= type of transmission

* drv: f = front-wheel drive, r = rear wheel drive, 4 = 4wd

* cty= city miles per gallon

* hwy= highway miles per gallon

* fl= fuel type

* class= "type" of car


```{r}
str(mpg)
summary(mpg)
```

### THE BASICS

In order to create a plot, you:
  
1.  Call the ggplot() function which creates a blank canvas
```{r}
ggplot(mpg)
```


2.  Specify aesthetic mappings, which specifies how you want to map variables to visual aspects. In this case we are simply mapping the displ and hwy variables to the x- and y-axes.
```{r}
ggplot(mpg, aes(x = displ, y = hwy))
```


3. You then add new layers that are geometric objects which will show up on the plot. In this case we add geom_point to add a layer with points (dot) elements as the geometric shapes to represent the data.
```{r}
ggplot(mpg, aes(x = displ, y = hwy)) +
  geom_point()
```


## ASTHEICS MAPPING

The aesthetic mappings take properties of the data and use them to influence visual characteristics,such as position, color, size, shape, or transparency.
Each visual characteristic can thus encode an aspect of the data and be used to convey information.

All aesthetics for a plot are specified in the aes() function call (later in this tutorial you will see that each geom layer can have its own aes specification).
For example, we can add a mapping from the class of the cars to a color characteristic:

```{r}
ggplot(mpg, aes(x = displ, y = hwy, color = class)) +
  geom_point()
```

equally

```{r}
ggplot(mpg, aes(x = displ, y = hwy)) +
  geom_point(aes(color = class))
```


mapping vs assignment
```{r}
ggplot(mpg, aes(x = displ, y = hwy)) +
  geom_point(color = "blue")
```


### GEOMETRY SHAPES

Left column: x and y mapping needed!
```{r}
ggplot(mpg, aes(x = displ, y = hwy)) +
  geom_point()
```



```{r}
ggplot(mpg, aes(x = displ, y = hwy)) +
  geom_smooth()
```


Right column: no y mapping needed!
```{r}
ggplot(data = mpg, aes(x = class)) +
  geom_bar()  
```


```{r}
ggplot(data = mpg, aes(x = hwy)) +
  geom_histogram() 
```


What makes this really powerful is that you can add multiple geometries to a plot, 
thus allowing you to create complex graphics showing multiple aspects of your data.

plot with both points and smoothed line
```{r}
ggplot(mpg, aes(x = displ, y = hwy)) +
  geom_point() +
  geom_smooth()
```


Of course the aesthetics for each geom can be different, so you could show multiple lines on the same plot (or with different colors, styles, etc). 
It’s also possible to give each geom a different data argument, so that you can show multiple data sets in the same plot.

For example, we can plot both points and a smoothed line for the same x and y variable 
but specify unique colors within each geom:
  
```{r}
ggplot(mpg, aes(x = displ, y = hwy)) +
  geom_point(color = "blue") +
  geom_smooth(color = "red")
```


So as you can see if we specify an aesthetic within ggplot it will be passed on to each geom that follows. 
Or we can specify certain aes within each geom, which allows us to only show 
certain characteristics for that specificy layer (i.e. geom_point).
  
color aesthetic passed to each geom layer
```{r}
ggplot(mpg, aes(x = displ, y = hwy, color = class)) +
    geom_point() +
    geom_smooth(se = FALSE) #se parameter: Display confidence interval around smooth?
```

color aesthetic specified for only the geom_point layer
```{r}
 ggplot(mpg, aes(x = displ, y = hwy)) +
    geom_point(aes(color = class)) +
    geom_smooth(se = FALSE)
```


### STATISTICAL TRANSFORMATION

If you look at the below bar chart, you’ll notice thatthe y axis was defined for us as the count of elements that have the particular type. 
This count isn’t part of the data set (it’s not a column in mpg), but is instead a statistical transformation that the geom_bar automatically applies to the data. 
In particular, it applies the stat_count transformation.

```{r}
ggplot(mpg, aes(x = class)) +
  geom_bar()
```


ggplot2 supports many different statistical transformations. 
For example, the “identity” transformation will leave the data “as is”. 
You can specify which statistical transformation a geom uses by passing it as the stat argument.
For example, consider our data already had the count as a variable,
We can use stat = "identity" within geom_bar to plot our bar height values to this variable.
Also, note that we now include y variable in ggplot function:
  
```{r}
  class_count <- dplyr::count(mpg, class)
  class_count
  
  ggplot(class_count, aes(x = class, y = n)) +
    geom_bar(stat = "identity")
```

  
We can also call stat_ functions directly to add additional layers. 
For example, here we create a scatter plot of highway miles for each displacement value and then use stat_summary to plot the mean highway miles at each displacement value.
  
```{r}
  ggplot(mpg, aes(displ, hwy)) + 
    geom_point(color = "yellowgreen") + 
    stat_summary(fun.y = "mean", geom = "line", size = 1, linetype = "dashed")
```


### POSITION ADJUSTEMENT

In addition to a default statistical transformation, each geom also has a default position adjustment which specifies a set of “rules” as to how different components should be positioned relative to each other. 
This position is noticeable in a geom_bar if you map a different variable to the color visual characteristic:
    
bar chart of class, colored by drive (front, rear, 4-wheel)
```{r}
ggplot(mpg, aes(x = class, fill = drv)) + 
  geom_bar()
```

  
The geom_bar by default uses a position adjustment of "stack", which makes each rectangle’s height proprotional to its value and stacks them on top of each other. 
We can use the position argument to specify what position adjustment rules to follow:
    
position = "dodge": values next to each other
```{r}
ggplot(mpg, aes(x = class, fill = drv)) + 
  geom_bar(position = "dodge")
```

position = "fill": percentage chart
```{r}
ggplot(mpg, aes(x = class, fill = drv)) + 
  geom_bar(position = "fill")
```


### MANAGING SCALES

Whenever you specify an aesthetic mapping, ggplot uses a particular scale to determine the range of values that the data should map to. 

color the data by engine type
```{r}
ggplot(mpg, aes(x = displ, y = hwy, color = class)) +
  geom_point()
```

same as above, with explicit scales
```{r}
ggplot(mpg, aes(x = displ, y = hwy, color = class)) +
  geom_point() +
  scale_x_continuous() +
  scale_y_continuous() +
  scale_colour_discrete()
```

Each scale can be represented by a function with the following name: scale_, followed by the name of the aesthetic property, followed by an _ and the name of the scale. 
A continuous scale will handle things like numeric data (where there is a continuous set of numbers), whereas a discrete scale will handle things like colors (since there is a small list of distinct colors).

While the default scales will work fine, it is possible to explicitly add different scales  to replace the defaults. For example, you can use a scale to change the direction of an axis:
  
```{r}
  ggplot(mpg, aes(x = cty, y = hwy)) +
  geom_point() +
  scale_x_reverse() +
  scale_y_reverse() 
```
  
Similarly, you can use scale_x_log10() and scale_x_sqrt() to transform your scale. 

```{r}
ggplot(mpg, aes(x = cty, y = hwy)) +
    geom_point() +
    scale_x_sqrt()
```
  
You can also use scales to format your axes:
```{r}
ggplot(mpg, aes(x = class, fill = drv)) + 
    geom_bar(position = "fill") +
    scale_y_continuous(breaks = seq(0, 1, by = .2), labels = scales::percent)
```
  

A common parameter to change is which set of colors to use in a plot. 
While you can use the default coloring, a more common option is to leverage the pre-defined palettes from colorbrewer.org. 
These color sets have been carefully designed to look good and to be viewable to people with certain forms of color blindness. 
We can leverage color brewer palletes by specifying the scale_color_brewer() function, passing the pallete as an argument.
  
default color brewer
```{r}
ggplot(mpg, aes(x = displ, y = hwy, color = class)) +
    geom_point() +
    scale_color_brewer()
```

specifying color palette
```{r}
ggplot(mpg, aes(x = displ, y = hwy, color = class)) +
    geom_point() +
    scale_color_brewer(palette = "Set3")
```

https://observablehq.com/@d3/color-schemes

You can also specify continuous color values by using a gradient scale, or manually specify the colors you want to use as a named vector.

```{r}
ggplot(mpg, aes(x = displ, y = hwy, color = class)) +
  geom_point() +
  scale_color_manual(values = c("yellowgreen", "orange","red","blue","gray","purple","violet"))

```

analogous for fill aes
```{r}
ggplot(mpg, aes(x = class, fill = drv)) + 
  geom_bar(position = "dodge")+
  scale_fill_brewer(palette = "Set1")
```


### COORDINATE SYSTEM

```{r}
ggplot(mpg, aes(x = class, fill = drv)) + 
  geom_bar(position = "dodge")+
  scale_fill_brewer(palette = "Set1")+
  coord_flip()
```


```{r}
ggplot(mpg, aes(x = class, fill = drv)) + 
  geom_bar(position = "dodge")+
  scale_fill_brewer(palette = "Set1")+
  coord_polar()
```


```{r}
ggplot(mpg, aes(x = class, fill = drv)) + 
  geom_bar(position = "dodge")+
  scale_fill_brewer(palette = "Set1")+
  coord_polar()

```

```{r}
ggplot(mpg, aes(x = "", fill = drv)) + 
  geom_bar()+
  scale_fill_brewer(palette = "Set1")+
  coord_polar("y")
```


### THEME

```{r}
ggplot(mpg, aes(x = "", fill = drv)) + 
  geom_bar()+
  scale_fill_brewer(palette = "Set1")+
  coord_polar("y")+
  theme_minimal()

```

```{r}
ggplot(mpg, aes(x = "", fill = drv)) + 
  geom_bar()+
  scale_fill_brewer(palette = "Set1")+
  coord_polar("y")+
  theme_void()
```


```{r}
ggplot(mpg, aes(x = "", fill = drv)) + 
  geom_bar()+
  scale_fill_brewer(palette = "Set1")+
  coord_polar("y")+
  theme_dark()
```


### FACETS

```{r}
ggplot(mpg, aes(x = displ, y = hwy, color=class)) +
  geom_point() 
```


```{r}
ggplot(mpg, aes(x = displ, y = hwy, color=class)) +
  geom_point() +
  facet_grid(~ class)
```

```{r}
ggplot(mpg, aes(x = displ, y = hwy)) +
  geom_point() +
  facet_grid(year ~ cyl)
```


### LABELS AND ANNOTATIONS

You can add titles and axis labels to a chart using the labs() function

```{r}
ggplot(mpg, aes(x = displ, y = hwy, color = class)) +
  geom_point() +
  labs(title = "Fuel Efficiency by Engine Power",
       subtitle = "Fuel economy data from 1999 and 2008 for 38 popular models of cars",
       x = "Engine power (litres displacement)",
       y = "Fuel Efficiency (miles per gallon)",
       color = "Car Type")
```

It is also possible to add labels into the plot itself (e.g., to label each point or line) by adding a new geom_text or geom_label to the plot; effectively, you’re plotting an extra set of data which happen to be the variable names:

a data table of each car that has best efficiency of its type
```{r}
best_in_class <- mpg %>%
  group_by(class) %>%
  filter(row_number(desc(hwy)) == 1)

ggplot(mpg, aes(x = displ, y = hwy)) + 
  geom_point(aes(color = class)) +
  geom_label(data = best_in_class, aes(label = model), alpha = 0.5)
```

